hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-28 inst/manuscript/paper.Rmd
fix typos, move lib call to setup chunk

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remian largely confined to salmonids, likelu due to the availability of existing parameterizations and the commercial and recreational importance of these species.

To extrend these modeling capabilities beyond salminds and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations. **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017. Users can also develop custom models using their own or published temperature and phenological data within the effective value framework.

To maximize accessibility, **hatchR** is available in two formats. The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing mutiple variables, such as phenology type, spawn timing, or thermal regimes. Comprehensize documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>). The second is a Shiny-based web application (@chang2024), offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality. Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.

```{r echo=FALSE, message=FALSE}
### 
library(tidyverse); library(hatchR); library(ggdist); library(ggridges)
```

>>>>

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(hatchR)
library(ggdist)
library(ggridges)
```

@@

This effective value approach has since been widely applied to salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.

To extend these modeling capabilities beyond Salmonids and facilitate broader applications, we developed **hatchR**, a software ecosystem designed to predict hatch and emergence timing for wild fish populations. **hatchR** enables users to input standard raw or summarized water temperature datasets commonly collected in field settings, conduct basic data validation, and apply built-in parameterizations such as those from @beacham1990 or @sparks2017. Users can also develop custom models using their own or published temperature and phenological data within the effective value framework.

To maximize accessibility, **hatchR** is available in two formats. The first is a R package, **hatchR**, distributed via CRAN (REF), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes. Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>). The second is a Shiny-based web application (@chang2024), offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality. Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.
